Thermodynamic response functions in a cell fluid model

### 2024-08-21 main_morse_response_functions.tex
remove not-needed beta
split eos into two subsections - in terms of temperature and chem potential, and another in terms of temperature and density

	\section{The equation of state}
	The equation of state obtained in~\cite[see Eq.(42)]{KozlovskiiDobush2020} reads
	\begin{equation}
		\label{eq:eosMT}
		Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
	\end{equation}
	The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

	The quantity $W(0)$ is expressed via parameters of the potential~(\ref{def:mod_morse}) as follows
	\begin{equation}
	W(0) = \Phi^{(r)}(0) \left[ B - 1 + \chi_0 + \tau (\chi_0 + A_\gamma) \right],
	\end{equation}
	where
	\begin{align*} 
		& B = 2 \gamma^3 e^{(1-\gamma)R_0/\alpha},
		\nonumber \\
		& A_\gamma = A e^{(n_0-\gamma)R_0/\alpha} \left( \gamma / n_0\right)^3, 
	\end{align*}
	and $\Phi^{(r)}(0)$ is the Fourier transform of the repulsive part of the potential at $\abs{\vb k}=0$
	\begin{equation*}
		\Phi^{(r)}(0) = \varepsilon C_H 8\pi {\rm e}^{\gamma R_0/\alpha} \left(\frac{\alpha}{\gamma}\right)^3.
	\end{equation*}

@@

	The coefficients $g_n$ are given by the formulas:
	\begin{align}
		& g_0 = \ln T_0, \qquad g_1 = T_1/T_0, \qquad g_2 = T_2/T_0 - g_1^2,  \nonumber \\
		& g_3 = T_3/T_0 - g_1^3 - 3g_1 g_2, \quad \qquad  g_4 = T_4/T_0 - g_1^4 - 6 g_1^2 g_2 - 4 g_1 g_3 - 3 g_2^2, \nonumber\\
		& a_4 = -g_4,
	\end{align}
	where $T_n(p,\alpha^*)$ are the following special functions
	\begin{equation}
		T_n(p,\alpha^*) = \sum_{m=0}^{\infty} \frac{(\alpha^*)^m}{m!} m^n {\rm e}^{-pm^2}.
	\end{equation}
	Here $\alpha^*=v e^{\beta_c\mu^*}$, and the parameter $p$ has the form
	\begin{equation}
		p = \frac{\beta_c}{2} \Phi^{(r)}(0) [\chi_0 + A_\gamma].
	\end{equation} 
	{\color{Red} {\bf{NOTE 1: }} Both quantities $p$ and $\alpha^*$ have the dimensions of volume. }

	The quantity $\beta_c$ denotes the critical value of the inverse temperature. In~\cite[see Eq.(31)]{KozlovskiiDobush2020} it was found that
	\begin{equation*}
		\varepsilon\beta_c = 0.200, \quad \frac{k_{\rm B} T_c}{\varepsilon} = 4.995.
	\end{equation*}
	Since $p$ is independent of temperature, the coefficients $g_n$ are also independent of temperature. The numerical values for other coefficients used in this paper are the same as those in~\cite[see Eqs.(5), (23), and (24)]{KozlovskiiDobush2020}:
	\begin{eqnarray*}
		\chi_0 = 0.07, & \quad \gamma = 1.65, \\
		n_0 = 1.521, & \quad R_0/\alpha = 2.9544, \\
		\alpha^* = 5.0 & \quad p = 1.0.
	\end{eqnarray*}

	The quantity $\tilde D(0)$ entering equations~\eqref{eq:eosMT} and~\eqref{chem_pot} is a function of temperature
	\begin{equation}
	\tilde D(0) = g_2 - \frac{1}{2} \frac{g_3^2}{g_4} - \frac{1}{\beta W(0)}.
	\end{equation}
	The condition $\tilde{D}(0) = 0$ defines the critical temperature~\cite[see Eq.(31)]{KozlovskiiDobush2020}
	\begin{equation}
		k_{\rm B}T_c = \left(g_2 - \frac{1}{2} \frac{g_3^2}{g_4} \right) (B - 1 + \chi_0) \Phi^{(r)}(0).
	\end{equation} 
	{\color{Red} {\bf{NOTE 2: }} In such representation, the critical temperature $k_{\rm B}T_c$ seems to have dimension of $[energy\times volume]$. }

@@

	........................................................

	The density $\eta$ of the many-particle system is the ratio of the average number of particles $\left\langle N \right\rangle$  to the total volume of the system $V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to the product $\beta \mu$. Using the general relation for the equation of state $PV\beta = \ln \Xi $ the density reads
	\begin{equation}
	\eta = \frac{\langle N\rangle}{v N_v} = \frac{1}{v} \frac{\partial}{\partial\beta\mu} Pv\beta = \frac{\bar n}{v},
	\end{equation} 
	where	
	\begin{equation}\label{eq:density}
	\bar n = n_g - \bar{M} + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{equation}\label{eq:crit_dens}
	n_g = g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2}.
	\end{equation}
	Combination of the equations \eqref{eq:density} and \eqref{eq:ro_M} reveals the explicit expression for the effective chemical potential $M$ as a function of the density and the temperature
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}
	where
	\begin{align} \label{eq:ro_nT}
		 & \rho_{n} = - 2 \left(\frac{g_3^2 - 2g_2 g_4}{g_4^2} \right)^{\frac{1}{2}} \cos \left( \frac{\alpha_n}{3} + \frac{\pi}{3} \right), \\
		 & \alpha_n = \arccos \left( \frac{3}{2} \left( - \frac{g_4^4}{\left( 2 g_2 g_4 - g_3^2\right)^3}\right)^{\frac{1}{2}} (n_g - \bar n)\right). \nonumber
	\end{align}

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
	\end{equation}
	Let us perform some transformations to rewrite the definition for $K_T$ into a form that is more suitable for the equation of state~\eqref{eq:eosMT}.
	\begin{eqnarray*}
		K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T,\langle N\rangle}}
		{\left(\partial P / \partial \mu\right)_{T,\langle N\rangle}},
	\end{eqnarray*}
	where $\rho = \langle N \rangle / V$ is the particle number density.
	Applying the Gibbs--Duhem equation at $T = const$, one has
	\begin{equation*}
		{\rm d} P = \rho {\rm d} \mu,
	\end{equation*}
	or
	\begin{equation}
		\label{eq:rho_dpdm}
		\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
	\end{equation}
	Substituting this into the last expression for $K_T$, one gets
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T,\langle N\rangle}
	\end{equation}
	Finally, from~\eqref{eq:rho_dpdm} it follows that
	\begin{equation}
		\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
	\end{equation}
	and ultimately we arrive at the very useful expression for the isothermal compressibility
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T,\langle N\rangle}
	\end{equation}

>>>>

	\section{The equation of state}
	\subsection{Pressure as a function of temperature and chemical potential}
	The equation of state obtained in~\cite[see Eq.(42)]{KozlovskiiDobush2020} reads
	\begin{equation}
		\label{eq:eosMT}
		Pv\beta = E_\mu(M, T) + M \bar \rho_0 + \frac{1}{2} \tilde D(0) \bar \rho_0^2 - \frac{a_4}{24} \bar \rho_0^4.
	\end{equation}
	The quantities in the left-hand side of the equation are $P$, the pressure; $\beta = (k_{\rm B} T)^{-1}$, the inverse temperature; $k_{\rm B}$, the Boltzmann constant; $T$, the temperature; $v$, cell volume. The quantities in the right-hand side are, in general, functions of the temperature $T$ and the chemical potential $\mu$. Let us present their expressions explicitly.

@@

	The quantity $W(0)$ is expressed via parameters of the potential~(\ref{def:mod_morse}) as follows
	\begin{equation}
	W(0) = \Phi^{(r)}(0) \left[ B - 1 + \chi_0 + \tau (\chi_0 + A_\gamma) \right],
	\end{equation}
	where
	\begin{align*} 
		& B = 2 \gamma^3 e^{(1-\gamma)R_0/\alpha},
		\nonumber \\
		& A_\gamma = A e^{(n_0-\gamma)R_0/\alpha} \left( \gamma / n_0\right)^3, 
	\end{align*}
	and $\Phi^{(r)}(0)$ is the Fourier transform of the repulsive part of the potential at $\abs{\vb k}=0$
	\begin{equation*}
		\Phi^{(r)}(0) = \varepsilon C_H 8\pi {\rm e}^{\gamma R_0/\alpha} \left(\frac{\alpha}{\gamma R_0}\right)^3.
	\end{equation*}

@@

	The coefficients $g_n$ are given by the formulas:
	\begin{align}
		& g_0 = \ln T_0, \qquad g_1 = T_1/T_0, \qquad g_2 = T_2/T_0 - g_1^2,  \nonumber \\
		& g_3 = T_3/T_0 - g_1^3 - 3g_1 g_2, \quad \qquad  g_4 = T_4/T_0 - g_1^4 - 6 g_1^2 g_2 - 4 g_1 g_3 - 3 g_2^2, \nonumber\\
		& a_4 = -g_4,
	\end{align}
	where $T_n(p,\alpha^*)$ are the following special functions
	\begin{equation}
		T_n(p,\alpha^*) = \sum_{m=0}^{\infty} \frac{(\alpha^*)^m}{m!} m^n {\rm e}^{-pm^2}.
	\end{equation}
	Here $\alpha^*=v e^{\beta_c\mu^*}$, and the parameter $p$ has the form
	\begin{equation}
		p = \frac{\beta_c}{2} \Phi^{(r)}(0) [\chi_0 + A_\gamma].
	\end{equation} 
	{\color{Red} {\bf{NOTE 1: }} Quantities $\alpha^*$ has the dimension of volume.}

	The quantity $\beta_c$ denotes the critical value of the inverse temperature. In~\cite[see Eq.(31)]{KozlovskiiDobush2020} it was found that
	\begin{equation*}
		\varepsilon\beta_c = 0.200, \quad \frac{k_{\rm B} T_c}{\varepsilon} = 4.995.
	\end{equation*}
	Since $p$ is independent of temperature, the coefficients $g_n$ are also independent of temperature. The numerical values for other coefficients used in this paper are the same as those in~\cite[see Eqs.(5), (23), and (24)]{KozlovskiiDobush2020}:
	\begin{eqnarray}
		\label{params}
		\chi_0 = 0.07, & \quad \gamma = 1.65, \\
		n_0 = 1.521, & \quad R_0/\alpha = 2.9544, \\
		\alpha^* = 5.0 & \quad p = 1.0.
	\end{eqnarray}

	The quantity $\tilde D(0)$ entering equations~\eqref{eq:eosMT} and~\eqref{chem_pot} is a function of temperature
	\begin{equation}
	\tilde D(0) = g_2 - \frac{1}{2} \frac{g_3^2}{g_4} - \frac{1}{\beta W(0)}.
	\end{equation}
	The condition $\tilde{D}(0) = 0$ defines the critical temperature~\cite[see Eq.(31)]{KozlovskiiDobush2020}
	\begin{equation}
		k_{\rm B}T_c = \left(g_2 - \frac{1}{2} \frac{g_3^2}{g_4} \right) (B - 1 + \chi_0) \Phi^{(r)}(0).
	\end{equation}

@@

	\subsection{Pressure as a function of temperature and density}
	In this work, by density we mean the particle number density $\rho = \langle N \rangle / V$.
	In the framework of the grand canonical ensemble the average number of particles $\left\langle N \right\rangle$ is a partial derivative of natural logarithm of the grand partition function $\Xi$ with respect to the product $\beta \mu$. 
	\begin{eqnarray*}
		\langle N \rangle & = & \left(\frac{\partial \ln\Xi}{\partial \beta \mu}\right)_{T,V} 
		\\
		& = & \left(\frac{\partial (PV)}{\partial \mu}\right)_{T,V}
		\\
		& = & V \left(\frac{\partial P}{\partial \mu}\right)_{T,V}.
	\end{eqnarray*}
	From this equation it follows
	\begin{equation*}
		\frac{\langle N \rangle}{V} = \frac{\langle N \rangle}{v N_v} = \left(\frac{\partial P}{\partial \mu}\right)_{T,V}
	\end{equation*}
	and thus
	\begin{equation}
		\bar{n} \equiv \frac{\langle N \rangle}{V} v = \frac{\langle N \rangle}{N_v} = v \left(\frac{\partial P}{\partial \mu}\right)_{T,V} 
		= \left(\frac{\partial (Pv\beta)}{\partial (\beta \mu)}\right)_{T,V}.
	\end{equation}
	The quantity $\bar{n}$, on the one hand, has the meaning of the reduced particle number density. On the other hand, it is the average number of particles per cell.
	Taking explicit derivatives, we arrive at
	\begin{equation}\label{eq:density}
	\bar n = n_g - M + \frac{ \bar \rho_0}{\beta W(0)}.
	\end{equation}
	The quantity $n_g$ in the equation \eqref{eq:density} is the critical density
	\begin{eqnarray}\label{eq:crit_dens}
	n_g & = & g_1 - \frac{g_3}{g_4}\left(  g_2 - \frac{1}{2} \frac{g_3^2}{g_4}\right) - \frac{1}{6} \frac{g_3^3}{g_4^2} 
	\nonumber\\
	& = & g_1 - \frac{g_2 g_3}{g_4} + \frac{g_3^3}{3g_4^2}
	\end{eqnarray}
	Its numerical value for parameters~\eqref{params} is
	\begin{equation*}
		n_g = 0.978.
	\end{equation*}
	Combination of the equations \eqref{eq:density} and \eqref{eq:ro_M} reveals the explicit expression for the effective chemical potential $M$ as a function of the density and the temperature
	\begin{equation}\label{eq:M_nT}
		\bar M = \frac{\rho_{n}}{\beta W(0)} - (\bar n - n_g),
	\end{equation}
	where
	\begin{align} \label{eq:ro_nT}
		 & \rho_{n} = - 2 \left(\frac{g_3^2 - 2g_2 g_4}{g_4^2} \right)^{\frac{1}{2}} \cos \left( \frac{\alpha_n}{3} + \frac{\pi}{3} \right), \\
		 & \alpha_n = \arccos \left( \frac{3}{2} \left( - \frac{g_4^4}{\left( 2 g_2 g_4 - g_3^2\right)^3}\right)^{\frac{1}{2}} (n_g - \bar n)\right). \nonumber
	\end{align}

@@

	\subsection{Isothermal compressibility}
	The isothermal compressibility is defined by
	\begin{equation}
		\label{def:isotherm_compres}
		K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
	\end{equation}
	Let us perform some transformations to rewrite the definition for $K_T$ into a form that is more suitable for the equation of state~\eqref{eq:eosMT}.
	\begin{eqnarray*}
		K_T & = & - \frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & -\rho \left(\frac{\partial \left(\frac{1}{\rho}\right)}{\partial P}\right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,\langle N\rangle}
		\\
		& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T,\langle N\rangle}}
		{\left(\partial P / \partial \mu\right)_{T,\langle N\rangle}},
	\end{eqnarray*}
	where $\rho = \langle N \rangle / V$ is the particle number density.
	Applying the Gibbs--Duhem equation
	\begin{equation}
		N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
	\end{equation}
	at $T = const$ one has 
	\begin{equation*}
		{\rm d} P = \rho {\rm d} \mu,
	\end{equation*}
	or
	\begin{equation}
		\label{eq:rho_dpdm}
		\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
	\end{equation}
	Substituting this into the last expression for $K_T$, one gets
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T,\langle N\rangle}
	\end{equation}
	Finally, from~\eqref{eq:rho_dpdm} it follows that
	\begin{equation}
		\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
	\end{equation}
	and ultimately we arrive at the very useful expression for the isothermal compressibility
	\begin{equation}
		K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T,\langle N\rangle}
	\end{equation}
